Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-04-17 analysis/paper/paper.Rmd
Set up Zotero web with RStudio

The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by studies employing more informally driven methods. This assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful. However, the approach taken is partially informed by research founded on the so-called Frison Effect. These scholars have argued that most morphological variation in retouched lithics is not the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part related to variable degrees of modification through use and rejuvination. Consequently, some artefact categories have here been collapased into a single category for the CA. This for example pertains to tool types such as scrapers, burins, drills and knives and indeterminate artefacts with retouch. That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn rendering their contribution as analytic units potentially entirely misleading. These have all been combined into the single category "small flint tools" here. (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material). While aggregating artefact groups in this manner could potentially subsume important variation, it does also reduce the possibility that any drawn conclusions are not simply the result of employing erroneous units of analysis. An underlying assumption is therefore that the retained categories represent artefacts categories that have fulfilled different purposes or are related to different technological processes. While ultimately intuitive in nature, it does seem reasonable to assume that large non-flint stone tools (axes, adzes, chisels, clubs and hatches) have fulfilled different purposes than for example the previously mentioned small flint tools.

However, the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for [@dibble2017], does have major implications that the above-outlined analysis does not take into account. For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life. While use-wear analysis could potentially offer a way to identify what artefacts were used for towards the end of their use-life, these kinds of analyses are extremely time-consuming and are therefore typically only conducted on a smaller number of artefacts that have already been selected for analysis based on their shape [e.g. @solheim2018]. Thus, while these analyses can potentially get at in-group variation among artefacts that we assume can be treated in aggregate, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017]. As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with these inferential limitations in mind. The logic behind these measures are founded in an understanding of lithic technology as being organised along a continuum ranging between curated and expedient [@binford1979]. An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of replacement. Curated technological organisation has been defined as the manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@bamforth1986]. The empirical correspondent to a curated technological organisation is consequently high degrees of retouch, as this is commonly seen as a means of extending the use-life of lithics by repeated rejuvination and modification of edges. The continuum between curated and expedient technology has in turn been related to settlement and mobility strategies, as well as raw-material quality and availability [@smith2015]. Higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability, and allow for the accumulation of material resources. Higher raw-material availability should in turn reduce the necessity for extensive retouch, meaning the material expectation for a lower degree of mobility is relatively high density of lithics and a low relative frequency of retouched artefacts. Additional empirical expectation for

@@

A note should also be made on the fact that a couple of variables that are sometimes invoked for the classification of sites in terms of associated mobility pattern are omitted here. This pertains to number of site features such as fireplaces, cooking pits and dwelling structures. The omission of feature count from the analysis follows from the fact that taphonomic loss is likely to lead to a chronological bias in the preservation and associated diffuclty of discovering of features. Similarly, the number of activity areas, effectively number artefact clusters, however defined, has also been disregarded. This follows most notably from the fact that the impact of post-depositional processes at Stone Age sites in Norway is arguably understudied. This pertains for example to the impact of bioturbation in the form of three-throws, which can can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here [@jorgensen2017; @darmark2018e].

>>>>

The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods. This assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful. However, the approach taken is partially informed by the so-called Frison Effect, which highlights the fact that the shape of lithics studied by archaeologists when they where discarded and could have had long and complex use-lifes in which they took on a multitude of different shapes. Starting with , several scholars have argued that morphological variation in retouched lithics is not predominantly the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvination [e.g. @dibble1995] . Consequently, some artefact categories have here been collapased for the CA. This for example pertains to tool types such as scrapers, burins, drills and knives and indeterminate artefacts with retouch. That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn rendering their contribution as analytic units potentially entirely misleading. These have all been combined into the single category "small flint tools". (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material). While aggregating artefact groups in this manner could potentially subsume important variation, it does also reduce the possibility that any drawn conclusions are not simply the result of employing erroneous units of analysis. An underlying assumption is therefore that the retained categories represent artefacts categories that have fulfilled different purposes or are related to different technological processes. While ultimately intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.

However, the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for does have major implications that the above-outlined analysis does not manage to take into account [@dibble2017]. This renders it difficult to align any of the results with specific behavioural dimensions. For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life. While use-wear analysis could potentially offer a way to identify what artefacts were used for towards the end of their use-life, these kinds of analyses are extremely time-consuming and are therefore typically only conducted on a smaller number of artefacts that have already been selected for analysis based on their shape [e.g. @solheim2018]. Thus, while these analyses can potentially get at in-group variation pertaining to the end state of artefacts that we assume can be treated in aggregate, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017]. As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with these inferential limitations in mind. The logic behind these measures are founded in an understanding of lithic technology as being organised along a continuum ranging between curated and expedient [@binford1979]. An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources for modification and rejuvination, resulting in high rates of replacement. Curated technological organisation, on the other hand, has been defined as the manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations [@binford1979, @bamforth1986]. The empirical correspondent to a curated technological organisation is consequently high degrees of retouch, as this is commonly seen as a means of extending the use-life of lithics by repeated rejuvination and modification of edges.

The continuum between curated and expedient technology has in turn been related to settlement and mobility strategies, as well as raw-material quality and availability [@andrefsky1994; @smith2015]. Higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability, and allow for the accumulation of material resources. Higher raw-material availability should in turn reduce the necessity for extensive retouch, meaning the material expectation for a lower degree of mobility is relatively high density of lithics and a low relative frequency of retouched artefacts. Additional empirical expectation for

@@

A note should also be made on the fact that a couple of variables that are sometimes invoked for the classification of sites in terms of associated mobility pattern are omitted here. This pertains to number of site features such as fireplaces, cooking pits and dwelling structures. The omission of feature count from the analysis follows from the fact that taphonomic loss is likely to lead to a chronological bias in the preservation and associated diffuclty of discovering of features. Similarly, the number of activity areas, effectively number artefact clusters, however defined, has also been disregarded. This follows most notably from the fact that the impact of post-depositional processes at Stone Age sites in Norway is arguably understudied. This pertains for example to the impact of bioturbation in the form of three-throws, which can can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here [@jørgensen2017; @darmark2018e].
